Supplementary Materials

### 2023-02-08 manuscript/manuscript.tex
ms updates

Previous work in population genomics has established that negative selection
against against new deleterious variants introduced by mutation profoundly
shapes large-scale patterns of variation across the genome. However, the
mathematical models that describe this process only consider variants so
deleterious that they cannot drift up in frequency and ultimately fix. In
reality, there number of mildly deleterious mutations

@@

@@

@@

\begin{figure}[!htb]
  \centering
  \includegraphics{figures/experimental_1.pdf}
  \caption{}
  \label{fig:figure-1}
\end{figure}

@@

@@

\begin{align}
  \label{eq:}
  {N}_{e}(x) &= N \exp \left( -\frac{V(x)}{2} Q^2(x) \right) & \text{\emph{effective population equation}} \\
  {R}(x) &= \frac{4N_e(x)U s}{\exp(4 N_e(x) s)-1}  & \text{\emph{ratchet equation}} 
\end{align}

@@

\begin{align}
  V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  Q^2(x) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us}{U-{R(x)}} & \text{\emph{}} \text{\emph{variance decay rate}}
\end{align}

and $U = \mu L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates for a segment $L$ basepairs long. Each mutation has a
selection coefficient of $s$, for which we assume additive effects. Then, the
local reduction factor is $b(x) = \nicefrac{N_e(x)}{N}$, and the per-basepair
per-generation substitution rate is $\lambda_D = \nicefrac{R(x)}{L}$.

In our software \texttt{bgspy}, these two equations are solved for a grid of
mutation rates and selection coefficients for each specified segment from a
feature type (e.g. CDS, UTRs, etc.). This results in estimates of $N_e(x)$ and
$R(x)$ per segment. In the first-round B' calculation, we use the
drift-effective population size, $N$.

@@

@@

@@

\begin{align}
  T \approx \frac{\exp(4 N_e s)-1}{2 U s N_e}.
\end{align}

Since the ratchet rate depends on $N_e$, which in turn depends on the rate of
the ratchet, we write the inverse as,

\begin{align}
  R(N_e) := \frac{2 U s N_e}{\exp(4 N_e s)-1}
\end{align}

@@

The solution to these equations is the asymptotic equilibrium $\widetilde{N}_e$ and
equilibrium ratchet rate $\widetilde{R}$. This set the equilibrium variance,
$\widetilde{V} = Us - 2\widetilde{R}s$.

>>>>

Previous work in population genomics has established that negative selection
against against new deleterious variants introduced by mutation profoundly
shapes large-scale patterns of variation across the genome. However, the
mathematical models that describe this process only consider variation so
deleterious that it cannot drift up in frequency and ultimately fix in the
population. In reality, many new mutations are only mildly deleterious, and can
fix in the population if random perturbations from genetic drift or selection
at neighboring sites (genetic draft) overpowers the selection against them.
Despite the large role negative selection plays across the human genome, no
existing evolutionary models consider the joint effects of mutation, selection,
and genetic drift and drift across the genome. Consequently, population genetic
models cannot describe the ultimate fate of every new variant ---eventual
extinction or fixation in the population--- and thus cannot connect to observed
rates of molecular evolution.

whole-genome variance models.

This is because, in the very long run, all mutations must be accounted for.

Here, we introduce a new whole-genome general mutation-selection-draft-drift
(MSD$^2$) model that extends and implements previous theory by
\textcite{Santiago2016-mu}.

Unlike classic mutation-selection models which cannot accurately capture the
dynamics of weakly deleterious mutation, MSD$^2$ models can, and consequently
can provide estimates of the distribution of selection coefficients of new
mutations. Additionally, because our model incorporates drift and draft, our
model makes explicit predictions about the substitution rate of weakly
deleterious mutations and allows us to test XXX.

@@

If the human evolutionary process were strongly impacted by Hill--Robertson
Interference, we should expect that substitution rates are highest in regions
experiencing the most selection, which are regions when $B(x)$ is the lowest.
However, our model

emphasis on whole genome models

@@

At every spot in the genome the processes of mutation-selection-draft-drift
against alleles deleterious in most environments reduce pairwise diversity. We
can imagine some unobserved reduction map, $B(x)$ across the genome. Here we
derive theory for this map under these processes and show how statistical
models can be fit using this theory.

@@

%\begin{figure}[!htb]
%  \centering
%  \includegraphics{figures/experimental_1.pdf}
%  \caption{}
%  \label{fig:figure-1}
%\end{figure}

@@

\section*{Discussion}

It is worth considering why this model performs well for models of negative
selection. The fitness variance equation, $V(x) = (U-2R)s$ tells us that under
MSD$^$2 models, the long-run variance is set by the balance of mutation and
fixation. Whether a particular basepair's behavior follows this equation
depends upon whether the long-run average is close to ``typical" for that
basepair at a given time. Good things fix so quickly, that we're left with a
constant churning of new variation

@@

\begin{align}
  \label{eq:}
  {N}_{e}(x) &= N \exp \left( -\frac{V(x)}{2} Q^2(x) \right) & \text{\emph{effective population equation}} \\
  {R}(x) &= \frac{4N_e(x)U s}{\exp(4 N_e(x) s)-1}  & \text{\emph{substitution equation}} 
\end{align}

@@

\begin{align}
  %V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  V(x) &= (U  - 2 {R}(x)) s & \text{\emph{fitness variance equation}} \\
  Q^2(x) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us}{U-{R(x)}} & \text{\emph{}} \text{\emph{variance decay rate}}
\end{align}

and $U = \mu L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates for a segment $L$ basepairs long. Each mutation has a
selection coefficient of $s$, for which we assume additive effects. Then, the
local reduction factor is $b(x) = \nicefrac{N_e(x)}{N}$, and the per-basepair
per-generation substitution rate is $\lambda_D = \nicefrac{R(x)}{L}$. Note that
in the variance equation, the long-run variance is the balance between mutation
and twice the loss due to substitution, $(U-2R)s$, since for selective balance
$V(x) = 0$ for $s>0$, the per-basepair substitution rate must be half the mutation rate,
since each substitution removes twice the variance introduced by a mutation in
a diploid species.

In our software \texttt{bgspy}, these two equations are solved for a grid of
mutation rates and selection coefficients for each specified segment from a
feature type (e.g. CDS, UTRs, etc.). This results in estimates of $N_e(x)$ and
$R(x)$ per segment. In the first-round B' calculation, we use a single
pre-specified drift-effective population size, $N$. The B' maps created here at
this step reflect selection only in the segment, ignoring the impact of
selection at other segments. By contrast, the locally-rescaled B' maps solve
these equations, assuming the drift-effective population size is $B(x)N$. These
reflect the segment's dynamics if they were effected by selection at other
segments under selection in the genome. Consequently, we can compare and
contrast models with local-recalling to test whether we achieve a better fit if
we account selective interference from other segments.

@@

\subsection*{Assessing the B' Maps}

We first assessed the B' maps against the observed average reduction across
$XXX$ forward-simulations, $\bar{B}_\text{sim}(x)$. The mean-squared error has
two sources, (1) bias from a breakdown in our theory $B(x)$ for the average
reduction, and (2) variance from the simulations. We show that for most of the
parameter space, the measured mean-squared error between simulation averages
and our B' maps is at the theoretic minimum. This minimum variance is entirely
from the coalescent noise when estimating $\bar{B}\text{sim}(x), which

@@

\subsection*{Substitution Rate Predictions}

Our MLE fit $\hat{\Phi}$ also implies a set of corresponding substitution rate
predictions for each segment under mutation-selection-draft process. There are
two sources of variation in these predictions in this model. First, there is
the variation across the pre-specified functional categories, e.g. CDS, UTRs,
and PhastCons. Second, there is region-specific variation

@@

\begin{align}
  T \approx \frac{\exp(4 N_e s)-1}{2 U s N_e}
\end{align}

Since the ratchet rate depends on $N_e$, which in turn depends on the rate of
the ratchet, we write the ratchet, or substitution, rate per basepair per
generation as,

\begin{align}
  R(N_e) := \frac{2 U s N_e}{\exp(4 N_e s)-1}.
\end{align}

@@

The solution to these equations is the asymptotic equilibrium $\widetilde{N}_e$
and equilibrium ratchet rate $\widetilde{R}$. This sets the equilibrium
variance, $\widetilde{V} = Us - 2\widetilde{R}s$, or $\widetilde{V} = (U -
2\widetilde{R})s$. Note that because $p_\text{fix} \le \nicefrac{1}{2N}$ for
deleterious to neutral variation, $2R \le U$, or the diploid mutation rate must
be greater than twice the per-basepair per-generation substitution rate for
equilibrium variance $\widetilde{V} > 0$ for $s>0$.

### 2023-02-08 manuscript/supplementary.tex
ms updates

  \item \textbf{Reference masking}: soft and hard-masked regions in the human GRCh38
    reference genome are also masked.

@@

These files are all produced by the Snakemake file \path{data/annotation/Snakefile}.

@@

where $h$ is the bandwidth or window size, $b$ the bin index for windows of
width $h$, $\mathcal{V}_b$ is the set of putatively neutral sites in bin $b$,
$\bar{\pi}(b | \Psi)$ are the average diversity in bin $b$, and
$Y_{\text{S},b}$ and $Y_{\text{S},b}$ are the sums across putatively neutral
sites within a bin. Note that by binning, we sum the pairwise summaries of the
data $\mathbf{Y}$ across sites, so the likelihood across bins is naturally
weighted by the quantity of observed data. We estimate $h$ by fitting the model
for a variety of window sizes $h$ and select the best using out-sample cross
validation, as is standard in non-parametric regression (XXX, Wasserman, p.
70).

@@

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g^S \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds
\end{align}

where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the reduction due to segment
$g$, and there are $S$ total segments. Here, $\mu(s, k(g))$ is the per-basepair
per-genome rate that mutations with selection coefficient $s$ enter segments
with feature class $k(g)$, $L_g$ and $r_g$ are the length in basepairs and
recombination rate per basepair of segment $g$, and the recombination distance
between focal site $x$ and segment $g$ is $d(x, g)$. The functional form of $b$
varies depending on whether the classic BGS model is used, or our form of
Santiago and Caballero's (\citeyear{Santiago2016-mu}) equations (see Section
XXX).

@@

\begin{align}
  b(\mu, s, L, r, 0) &= \exp \left( - \int_0^L \frac{2\mu(l)}{s (1 + (1-s) r(l)/s)^2} dl \right) 
\end{align}

@@

\begin{align}
  b(\mu, s, L, r_\text{BP}, 0) &= \exp \left( - \frac{2\mu L}{s + (1-s) r_\text{BP} L} \right)
\end{align}

@@

\begin{align}
  b(\mu, s, L, r_\text{BP}, d) &= \exp \left( \frac{-2\mu L}{(d(1-s) + s)(s + (d + r_\text{BP}L)(1-s))} \right)
\end{align}

@@

\begin{align}
  \widetilde{b}'(\mu, s, L, r_\text{BP}) = \frac{\widetilde{N}_{e,\infty}}{N}
\end{align}

@@

\begin{align}
  {N}_{e,\infty} &= N \exp \left( -\frac{V}{2} Q_\infty^2 \right) \\
  {R} &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
\end{align}

@@

Note that the asymptotic model above determines the reduction in effective
population size experienced by a focal site in the middle of a $L$-basepair
segment, but B maps require the reduction $b'$ experienced by a focal site $x$
at an arbitrary recombination distance apart. For each segment, we pre-compute
the solution to the system of two equations, giving us $V$ and $Q_\infty^2$,
and then use these values in the equation,

\begin{align}
  b'(\mu, s, L, r, d) = \exp\left( -\frac{V}{2} \left(\frac{1}{(1-(1-d)Z}\right)^2 \right) 
\end{align}

where $Z = 1 - Us / (U-\widetilde{R})$ and $V = Us - 2\widetilde{R}s$.

@@

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g \sum_{i} b(\mu_{i, \;k(g)}, s_i, L_g, r_g, d(x, g))
\end{align}

>>>>

  \item \textbf{Reference masking}: soft and hard-masked regions in the human
      GRCh38 reference genome are also masked. Soft-masked regions were
      determined by Ensembl (XXX), which uses Repeat Masker (XXX) and Dust
      (XXX).

@@

These files are all produced by the Snakemake file \path{data/annotation/Snakefile}.
Note that the levels of accessible putatively neutral bases

@@

where $h$ is the bandwidth or window size, $b$ the bin index for windows of
width $h$, $\mathcal{V}_b$ is the set of putatively neutral sites in bin $b$,
$\bar{\pi}(b | \Psi)$ are the average diversity in bin $b$, and
$Y_{\text{S},b}$ and $Y_{\text{S},b}$ are the sums across putatively neutral
sites within a bin. Note that by binning, we sum the pairwise summaries of the
data $\mathbf{Y}$ across sites, so the likelihood across bins is naturally
weighted by the quantity of observed data.

% TODO
%We estimate $h$ by fitting the model
%for a variety of window sizes $h$ and select the best using out-sample cross
%validation, as is standard in non-parametric regression (XXX, Wasserman, p.
%70).

@@

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) &= \prod_g^S \exp \left( \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right) \\
             &= \exp \left( \sum_g^S \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right)
\end{align}

where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the log reduction due to
segment $g$, and there are $S$ total segments. Here, $\mu(s, k(g))$ is the
per-basepair per-genome rate that mutations with selection coefficient $s$
enter segments with feature class $k(g)$, $L_g$ and $r_g$ are the length in
basepairs and recombination rate per basepair of segment $g$, and the
recombination distance between focal site $x$ and segment $g$ is $d(x, g)$. The
functional form of $b$ varies depending on whether the classic BGS model is
used, or our form of Santiago and Caballero's (\citeyear{Santiago2016-mu})
equations (see Section XXX).

@@

\begin{align}
  b(\mu, s, L, r, 0) &=  - \int_0^L \frac{2\mu(l)}{s (1 + (1-s) r(l)/s)^2} dl 
\end{align}

@@

\begin{align}
  b(\mu, s, L, r_\text{BP}, 0) &= - \frac{2\mu L}{s + (1-s) r_\text{BP} L} 
\end{align}

@@

\begin{align}
  b(\mu, s, L, r_\text{BP}, d) &= \frac{-2\mu L}{(d(1-s) + s)(s + (d + r_\text{BP}L)(1-s))}
\end{align}

@@

\begin{align}
  \widetilde{b}'(\mu, s, L, r_\text{BP}) = \log\left(\frac{\widetilde{N}_{e,\infty}}{N} \right)
\end{align}

@@

\begin{align}
    {N}_{e,\infty} &= N \exp \left( -\frac{V}{2} Q_\infty^2 \right) \label{eqn:Ne_VQ2} \\
  {R} &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1} \nonumber \\
\end{align}

@@

Note that the asymptotic model above determines the reduction in effective
population size experienced by a focal site in the middle of a $L$-basepair
segment, but B maps require the reduction $b'$ experienced by a focal site $x$
at an arbitrary recombination distance apart. For each segment, we pre-compute
the solution to the system of two equations, giving us $V$ and $Q_\infty^2$,
and then use these values in the equation \eqref{eqn:Ne_VQ2},

\begin{align} 
    b'(\mu, s, L, r, d) = -\frac{V}{2} \left(\frac{1}{(1-(1-d)Z}\right)^2 
\end{align}

where $Z = 1 - Us / (U-\widetilde{R})$ and $V = Us - 2\widetilde{R}s$. TODO
check d.

@@

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \exp \left( \sum_g^S \sum_{i} b\left( \mu_{i, \;k(g)}, s_i, L_g, r_g, d(x, g) \right)\right)
\end{align}

%TODO double check
